---
title: "scTME: Automated Tumor Microenvironment Annotation"
author: "Michael Januszyk"
date: "2026-08-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scTME: Automated Tumor Microenvironment Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`scTME` transfers cell type annotations from a pan-cancer tumor microenvironment
atlas onto a new single-cell dataset. The atlas is built into the package, so no
reference needs to be supplied. Annotations are available at two resolutions: 7
broad parent populations, or 45 transcriptionally distinct subpopulations.

This vignette annotates a small public dataset and compares the results against
that dataset's own published labels.

```{r install-packages, eval = FALSE, include = FALSE}
## Install Required Packages

# 1. Check for and install BiocManager from CRAN
if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

# 2. Check for and install SingleR from Bioconductor
if (!require("SingleR", quietly = TRUE)) {
    BiocManager::install("SingleR")
}

# 3. Check for and install SingleCellExperiment from Bioconductor
if (!require("SingleCellExperiment", quietly = TRUE)) {
    BiocManager::install("SingleCellExperiment")
}

# 4. Check for and install devtools from CRAN
if (!require("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

# 5. Check for and install scTME from GitHub
if (!require("scTME", quietly = TRUE)) {
    devtools::install_github("michaeljanuszyk/scTME")
}
```


```{r setup, message = FALSE, warning = FALSE}
## Load Packages
library(Seurat)
library(SingleR)
library(SingleCellExperiment)
library(scTME)
```

```{r, echo=FALSE}
if (requireNamespace("xaringanExtra", quietly = TRUE)) {
  xaringanExtra::use_clipboard(
    button_text = "&#128203; Copy",
    success_text = "&#10004; Copied!"
  )
}
```

## The reference atlas and the query dataset

The reference atlas loads from the package itself:

```{r}
data("scTME_atlas", package = "scTME")
```

The query is a heavily down-sampled (1,000 cell) subset of the Nieto 2021
immune cell atlas (*A single-cell tumor immune atlas for precision oncology*),
which ships with the package:

```{r}
query_path <- system.file("extdata", "nieto_2021_downsampled1k.rds", package = "scTME")
query_dataset <- readRDS(query_path)
```

A larger 10,000-cell version of the same dataset — the one used in the
accompanying manuscript — can be downloaded from
<https://tmeatlas.stanford.edu> under Analyze Data → Download. It gives higher
fidelity than the 1,000-cell subset used here, at the cost of runtime.

Note that this is a Seurat object. SingleCellExperiment objects can also be
passed to `scTME`.

```{r}
names(query_dataset@meta.data)
```

The Nieto dataset carries two levels of its own annotations (`lv1_annot` and
`lv2_annot`), which are the final assignments made in that study. We have also
added a `cluster` column, which indexes the sorted order of the `lv2_annot`
values. Using it as the grouping variable means the cluster-level examples below
test label transfer given correct groupings, rather than testing de novo
clustering; the annotation itself is still made without reference to the
published labels.

```{r}
table(query_dataset$lv1_annot)
```


```{r}
table(query_dataset$lv2_annot)
```


```{r}
table(query_dataset$cluster)
```

## Run scTME on the query dataset

Now we can go ahead and run `scTME`. This function accepts up to 7 parameters:

* **`obj`**: The Seurat or SingleCellExperiment query dataset on which to run the predictions.
* **`ref`**: (Default: `NULL`) The reference object powering scTME (`scTME_atlas`). If left as `NULL`, the package's built-in reference atlas will load automatically.
* **`clusters`**: (Default: `NULL`) If `NULL`, the function evaluates each cell independently. Alternatively, provide a character array or column name labeling the groups/clusters in the query dataset (e.g., `query_dataset$cluster` or `"cluster"`) to run faster, cluster-level predictions.
* **`level`**: (Default: `2`) Evaluates the query against all 45 high-resolution subpopulations. It can optionally be set to `1` to evaluate against the 7 broader parent populations.
* **`label.col`**: (Default: `NULL`) The specific metadata column in the reference object containing the cell type labels. If `NULL`, the function automatically selects the correct column based on your chosen `level`.
* **`max.cells`**: (Default: `50000`) To improve runtime, this value automatically downsamples the *query dataset* if it exceeds this threshold.
* **`return.details`**: (Default: `FALSE`) By default, `scTME` returns a vector of the most likely cell type predictions. If set to `TRUE`, the program returns a full list object containing additional prediction strength and confidence metrics.

Note that the query downsampling controlled by `max.cells` is random. Call
`set.seed()` beforehand if you need a reproducible result.

### Broad populations, cluster-level

```{r}
res = scTME( obj = query_dataset, ref = NULL, clusters = query_dataset$cluster, level=1, return.details = F )
```

```{r}
query_dataset$prediction <- unname(res)
```

We can now compare these predictions against those originally assigned by
Nieto. `scTME_generateHeatmap()` cross-tabulates the two, coloring by row
proportion and printing the raw cell counts.

```{r, fig.width=10, fig.height=8}
scTME_generateHeatmap(query_dataset,
                      pred_col  = "prediction",
                      annot_col = "lv1_annot")
```


### Subpopulations, cluster-level

Level 2 gives more precise predictions for each `query_dataset$cluster`.

```{r}
res = scTME( query_dataset, ref=NULL, clusters = query_dataset$cluster, level=2, return.details = F )

query_dataset$prediction = unname(res)
```


```{r, fig.width=10, fig.height=8}
scTME_generateHeatmap(query_dataset,
                      pred_col  = "prediction",
                      annot_col = "lv2_annot")
```


### Subpopulations, cell-level

Setting `clusters = NULL` annotates each cell independently. This is slower and
noisier than the cluster-level call above, but requires no prior grouping of the
query dataset.

```{r}
res = scTME( query_dataset, ref=NULL, clusters = NULL, level=2, return.details = F )
query_dataset$prediction = unname(res)
```


```{r, fig.width=10, fig.height=8}
scTME_generateHeatmap(query_dataset,
                      pred_col  = "prediction",
                      annot_col = "lv2_annot")
```


## Visualize Predictions on UMAP

To see how the predicted cell types map onto the transcriptomic landscape, we can visualize the cells in reduced dimensional space. Here, we project both the original high-resolution annotations and the new `scTME` predictions onto the UMAP embeddings using our custom plotting wrapper.

```{r plot-umap, fig.width=12, fig.height=5, warning=FALSE, message=FALSE}
# Plot the original annotations using the safe wrapper
plot_true <- plotPredictionUmap(query_dataset, reduction = "umap", pred_col = "lv2_annot")

# Plot the scTME predictions using the safe wrapper
plot_pred <- plotPredictionUmap(query_dataset, reduction = "umap", pred_col = "prediction")

# Only attempt to modify and combine the plots if the UMAP actually exists
if (!is.null(plot_true) && !is.null(plot_pred) &&
    requireNamespace("patchwork", quietly = TRUE)) {

  library(patchwork)

  # Override the default titles and remove legends for clean patchwork display
  plot_true <- plot_true + ggplot2::ggtitle("True Annotations") + Seurat::NoLegend()
  plot_pred <- plot_pred + ggplot2::ggtitle("scTME Predictions") + Seurat::NoLegend()

  # Display side-by-side
  plot_true + plot_pred

} else {
  message("Skipping UMAP visualization because the 'umap' reduction is missing from the object.")
}
```
