---
title: "scTME: Automated Tumor Microenvironment Annotation"
author: "Michael Januszyk"
date: "2026-08-17"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scTME: Automated Tumor Microenvironment Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r install-packages, eval = FALSE, include = FALSE}
## Install Required Packages

# 1. Check for and install BiocManager from CRAN
if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

# 2. Check for and install SingleR from Bioconductor
if (!require("SingleR", quietly = TRUE)) {
    BiocManager::install("SingleR")
}

# 3. Check for and install scrapper from Bioconductor
if (!require("scrapper", quietly = TRUE)) {
    BiocManager::install("scrapper")
}

# 4. Check for and install devtools from CRAN
if (!require("devtools", quietly = TRUE)) {
    install.packages("devtools")
}

# 5. Check for and install SingleCellExperiment from Bioconductor
if (!require("SingleCellExperiment", quietly = TRUE)) {
    BiocManager::install("SingleCellExperiment")
}

# 6. Check for and install scTME from GitHub
if (!require("scTME", quietly = TRUE)) {
    devtools::install_github("michaeljanuszyk/scTME")
}
```


```{r setup, message = FALSE, warning = FALSE}
## Load Packages
library(Seurat)
library(SingleR)
library(scrapper)
library(SingleCellExperiment)
library(scTME)
```

```{r, echo=FALSE}
xaringanExtra::use_clipboard(
  button_text = "&#128203; Copy",
  success_text = "&#10004; Copied!"
)
```

Then download a heavily down-sampled (1,000 total cells) from the Nieto
2021 immune cell atlas (A single-cell tumor immune atlas for precision
oncology), made available at <https://tmeatlas.stanford.edu> -->
Analyze Data --> Download --> nieto_2021_downsampled1k.rds There is
also a 10k cells object available here, which was used for the
accompanying manuscript, but we'll use the smaller one now to reduce
processing time.

```{r}
data("scTME_atlas", package = "scTME")
```

```{r}
query_path <- system.file("extdata", "nieto_2021_downsampled1k.rds", package = "scTME")
query_dataset <- readRDS(query_path)
```

Note that this is a Seurat object. SingleCellExperiment objects can also be passed to scTME.

```{r}
names(query_dataset@meta.data)
```

We can see that the Nieto dataset has two levels of annotations
(lv1_annot and lv2_annot), which are the final assignments made in that
group's study. We have also added a "cluster" parameter which represents
the sorted order of the lv2_annot values. This allows us to blindly
evaluate cells for this short vignette.

```{r}
table(query_dataset$lv1_annot)
```


```{r}
table(query_dataset$lv2_annot)
```


```{r}
table(query_dataset$cluster)
```

## Run scTME on the query dataset

Now we can go ahead and run `scTME`. This function accepts up to 8 parameters:

* **`obj`**: The Seurat or SingleCellExperiment query dataset on which to run the predictions.
* **`ref`**: (Default: `NULL`) The reference object powering scTME (`scTME_atlas`). If left as `NULL`, the package's built-in reference atlas will load automatically.
* **`clusters`**: (Default: `NULL`) If `NULL`, the function evaluates each cell independently. Alternatively, provide a character array or column name labeling the groups/clusters in the query dataset (e.g., `query_dataset$cluster` or `"cluster"`) to run faster, cluster-level predictions.
* **`level`**: (Default: `2`) Evaluates the query against all 50 high-resolution subpopulations. It can optionally be set to `1` to evaluate against the 8 broader parent populations.
* **`label.col`**: (Default: `NULL`) The specific metadata column in the reference object containing the cell type labels. If `NULL`, the function automatically selects the correct column based on your chosen `level`.
* **`max.cells`**: (Default: `50000`) To improve runtime, this value automatically downsamples the *query dataset* if it exceeds this threshold.
* **`fast`**: (Default: `FALSE`) If set to `TRUE`, this automatically downsamples the internal *scTME reference atlas* from 50,000 cells to 5,000 cells to significantly speed up computation time. 
* **`return.details`**: (Default: `FALSE`) By default, `scTME` returns a vector of the most likely cell type predictions. If set to `TRUE`, the program returns a full list object containing additional prediction strength and confidence metrics.


```{r}
res = scTME( obj = query_dataset, ref = NULL, clusters = query_dataset$cluster, level=1, return.details = F )
```

```{r}
query_dataset$prediction <- unname(res)
```

## Compare annotations

We can now, for example, compare these predictions from this atlas against those originally assigned by Nieto. (Using the 10k cells object results in higher fidelity than the 1k object)

```{r, fig.width=10, fig.height=8}
library(pheatmap)

# 1. Create the base matrix
conf_table <- table(query_dataset$prediction, query_dataset$lv1_annot)
conf_matrix <- as.matrix(conf_table)

# 2. Row Normalization (convert to proportions from 0.0 to 1.0)
norm_matrix <- conf_matrix / rowSums(conf_matrix)

# 3. Create a custom palette (Pure white for 0, then noticeable pink to dark red)
my_colors <- c("white", colorRampPalette(c("lightpink", "firebrick3"))(49))

# 4. Generate the heatmap
pheatmap(
  mat = norm_matrix, 
  display_numbers = conf_matrix, # Colors by proportion, but prints the RAW cell counts!
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = my_colors,
  main = "scTME Predictions vs. True Annotations"
)
```


## Compare annotations

We can also use the level 2 annotations to get more precise predictions for each query_dataset$cluster

```{r}
res = scTME( query_dataset, ref=NULL, clusters = query_dataset$cluster, level=2, return.details = F )

query_dataset$prediction = unname(res)
```


```{r, fig.width=10, fig.height=8}
library(pheatmap)

# 1. Create the base matrix
conf_table <- table(query_dataset$prediction, query_dataset$lv2_annot)
conf_matrix <- as.matrix(conf_table)

# 2. Row Normalization (convert to proportions from 0.0 to 1.0)
norm_matrix <- conf_matrix / rowSums(conf_matrix)

# 3. Create a custom palette (Pure white for 0, then noticeable pink to dark red)
my_colors <- c("white", colorRampPalette(c("lightpink", "firebrick3"))(49))

# 4. Generate the heatmap
pheatmap(
  mat = norm_matrix, 
  display_numbers = conf_matrix, # Colors by proportion, but prints the RAW cell counts!
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = my_colors,
  main = "scTME Predictions vs. True Annotations"
)
```


```{r}
res = scTME( query_dataset, ref=NULL, clusters = NULL, level=2, return.details = F )
query_dataset$prediction = unname(res)
```


```{r, fig.width=10, fig.height=8}
library(pheatmap)

# 1. Create the base matrix
conf_table <- table(query_dataset$prediction, query_dataset$lv2_annot)
conf_matrix <- as.matrix(conf_table)

# 2. Row Normalization (convert to proportions from 0.0 to 1.0)
norm_matrix <- conf_matrix / rowSums(conf_matrix)

# 3. Create a custom palette (Pure white for 0, then noticeable pink to dark red)
my_colors <- c("white", colorRampPalette(c("lightpink", "firebrick3"))(49))

# 4. Generate the heatmap
pheatmap(
  mat = norm_matrix, 
  display_numbers = conf_matrix, # Colors by proportion, but prints the RAW cell counts!
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = my_colors,
  main = "scTME Predictions vs. True Annotations"
)
```


## Visualize Predictions on UMAP

#To see how the predicted cell types map onto the transcriptomic landscape, we can visualize the cells in reduced dimensional space. Here, we project both the original high-resolution annotations and the new `scTME` predictions onto the UMAP embeddings using Seurat's `DimPlot` function.


```{r plot-umap, fig.width=12, fig.height=5, warning=FALSE, message=FALSE}
library(patchwork)

# Plot the original annotations
plot_true <- DimPlot(query_dataset, reduction = "umap", group.by = "lv2_annot", label = TRUE, repel = TRUE) + 
  ggplot2::ggtitle("True Annotations") +
  NoLegend()

# Plot the scTME predictions
plot_pred <- DimPlot(query_dataset, reduction = "umap", group.by = "prediction", label = TRUE, repel = TRUE) + 
  ggplot2::ggtitle("scTME Predictions") +
  NoLegend()

# Display side-by-side using patchwork
plot_true + plot_pred
```

